---
title: "pov_full_emp"
author: "Daniel Perez"
date: "10/11/2021"
output: html_document
---

```{r libraries, include=FALSE}
library(tidyverse)
library(data.table)
library(epiextractr)
library(epidatatools)
library(labelled)
library(here)
library(blsAPI)
library(openxlsx)
library(zoo)
library(ggplot2)
```

Our benchmark 
https://www.cbpp.org/research/poverty-and-inequality/research-note-number-of-people-in-families-with-below-poverty#_ftnref2

[1] These figures reflect three refinements to a previous version of this analysis released on June 16, 2020. 

First, when calculating weekly earnings, we assigned zero earnings to workers who reported that they were absent from work all week without pay, even if they reported a “usual weekly earnings” amount. 
Second, for individuals classified as hourly workers, we calculated their weekly earnings as their hourly wage times their actual hours worked during the reference week. 
Third, we excluded families where all individuals who are employed (or who were employed recently) are self-employed because Census does not report weekly self-employment earnings.

```{r Load poverty thresholds and CPI}

annual_cpi <- read_csv(here('input/cpi_annual.csv'), col_names=TRUE)

cpi <- read_csv(here('input/cpi_monthly.csv'), col_names=TRUE) %>% 
  select(year, month, cpi_u, cpiurs)

cpi_2020 <- annual_cpi$cpi_u[annual_cpi$year=='2020']

cpi_base <- cpi$cpiurs[cpi$year == '2020' & cpi$month=='6'] # Adjust all to June 2020 $


pov_thresh <- read_csv(here('input/povthresh20.csv'), col_names=TRUE) %>%
  filter(over65==0) %>% 
  select(year, month, famsize, child_under18, pov_thresh_2020)

#create a historic threshold from 1994 to present
historic_thresh <- cpi %>%
  filter(year %in% c(1994:2020)) %>% 
  left_join(pov_thresh, by='month') %>%
  mutate(rpov_thresh = pov_thresh_2020*(cpi_u/cpi_2020)) %>% 
  rename(year = year.x) %>% 
  select(-year.y)

```


```{r Load data}

cps <- data.table(load_basic(1994:2020, orgwgt, basicwgt, finalwgt, year, month, minsamp, selfemp, selfinc, personid, hhid, hhtype, famtype, famid, famrel, ownchild, agechild, paidhre, wage, wageotc, weekpay, hourslwt, hourslw1, hoursu1i, emphrs, emp, unemp, cow1, lfstat, age, wbhao, wbho_only, faminc)) %>% 
  #keep only those interviewed in the ORG
  filter(minsamp %in% c(4,8), age<65) %>%
  mutate(selfemp0 = ifelse(selfemp==1 & !is.na(selfemp), yes=1, no=0),
         selfinc0 = ifelse(selfinc==1 & !is.na(selfinc), yes=1, no=0),
         selfany = ifelse(selfinc0==1 | selfemp0==1, yes=1, no=0),
         emp0 = ifelse(emp==1 & !is.na(emp), yes=1, no=0)) %>%
  #Create indicator for workers absent from work
  mutate(hourslw_i = ifelse(hourslwt<=0 | is.na(hourslwt), yes=0, no=1)) %>% 
  #Merge CPI data
  left_join(cpi, by=c('year', 'month')) %>% 
  #create indicator of reason not at work/hours last week
  mutate(realweekpay = weekpay*(cpiurs/cpi_base),
         realwageotc = wageotc*(cpiurs/cpi_base),
         realwage.c = ifelse(hoursu1i>0, yes=hoursu1i * realwageotc, no = NA),
         
         #If worker is paid hourly, calculate weekly pay as realwageotc*hoursu1i. 
         #If not hourly then weekpay== inflation adjusted CPS weekly pay variable.
         realweekpay.c = ifelse(paidhre==1, yes=realwage.c, no=realweekpay),
         
         #Assign workers with zero hours last week 0 wages
         realweekpay.c = replace(realweekpay.c, hourslw_i==0, NA),

         #Create annual pay variables
         realyearpay.c = ifelse(realweekpay.c>0, realweekpay.c * 52, no=NA),
         
         #create indicators for positive hours, weekly pay, and wages
         under18 = ifelse(age<18, yes=1, no=0)) %>% 
  set_value_labels(under18 = c('under 18'= 1, 'over 18'=0),
                   month = c('Jan' = 1, 'Feb' = 2, 'Mar' = 3, 'Apr' = 4,
                             'May' = 5, 'Jun' = 6, 'Jul' = 7, 'Aug' = 8,
                             'Sept' = 9, 'Oct' = 10, 'Nov' = 11, 'Dec' = 12),
                   emp0 = c('Employed' = 1, 'NILF/unemployed/NIU'=0)) %>% 
  group_by(year, month, hhid, famid) %>% 
  #create group id for each family in sample
  mutate(group_id = cur_group_id()) %>% 
  ungroup()

#Create weights for children under 16
wgt_ratios <- cps %>% 
  select(group_id, famrel, orgwgt, finalwgt) %>% 
  filter(famrel==1) %>% 
  group_by(group_id) %>% 
  #Calculate ratio of ORG:final weight according to Saenz Sherman 2020.
  mutate(orgfinal_ratio = ifelse(famrel==1, yes=orgwgt/finalwgt, no=NA)) %>% 
  select(group_id, orgfinal_ratio)

```



```{r Clean CPS Basic}

thedata <- cps %>% 
  #merge org-to-final weight ratios to cps dataset by group_id
  left_join(wgt_ratios, by='group_id') %>% 
  relocate(famrel, age, finalwgt, orgwgt, .after=orgfinal_ratio) %>% 
  #Create a new weight for those under 16yo according to Saenz, Sherman 2020.
  mutate(adj_wgt = ifelse(age<16, yes=orgfinal_ratio*finalwgt, no=orgwgt))

#Create new data frame to measure total family income, excluding self-employed, self-incorporated families
fam_incomes <- thedata %>% 
  group_by(group_id) %>%
    #This section removes families where all working adults are self-employed
  mutate(selfemp_pop = sum(selfany==1), #count self employed in each family
         emp_pop = sum(emp0==1, na.rm=TRUE)) %>%
  #remove families where all workers are self-employed
  filter(!(emp_pop<=selfemp_pop & emp_pop!=0)) %>% 
  #poverty thresholds count children in family under 18
  mutate(famsize = n(),
         child_under18 = sum(famrel==3 & age<18)) %>% 
  #count family's real annual pay
  mutate(rtotweekpay = sum(realweekpay.c, na.rm=TRUE),
         rtotyearpay = sum(realweekpay.c*52, na.rm=TRUE)) %>% 
  ungroup()

fam_poverty <- fam_incomes %>% 
  #Merge poverty thresholds from census
  left_join(historic_thresh) %>% 
  mutate(rweekpov_thresh = (rpov_thresh/52)) %>%
  #create unemp variable
  mutate(unemp = ifelse(lfstat==2, yes=1, no=0)) %>%
  set_value_labels(unemp = c('Unemployed'=1, 'Employed/NILF'=0)) %>% 
  mutate(inpoverty = ifelse((rtotweekpay<rweekpov_thresh), yes=1, no=0), #Create an indicator for folks in poverty
         povlevel = round((rtotyearpay/rpov_thresh)*100)) %>% #create poverty measure
  mutate(noincome = ifelse(rtotweekpay==0, yes=1, no=0)) %>% #count folks w/no income
  set_value_labels(noincome = c('No income'= 1, 'With income'= 0)) %>% 
  #remove 1152 observations in sample that have no ORG weight
  filter(!is.na(adj_wgt))

```

```{r Time series of monthly poverty}

natl_poverty <- fam_poverty %>% 
  group_by(year, month) %>% 
  summarize(n = n(),
            in_poverty = sum(inpoverty * (adj_wgt/12), na.rm=TRUE),
            poverty = weighted.mean(inpoverty, w=adj_wgt, na.rm=TRUE),
            urate = weighted.mean(unemp, w=adj_wgt, na.rm=TRUE)) %>%
  mutate(group='Poverty, all',
         poverty_earnings='All',
         date= as.Date(paste(year, month, 1, sep = "-"), "%Y-%m-%d"))
  
    
hist_poverty <- ggplot(aes(x=date, y=poverty), data=natl_poverty)+
  theme_light()+
  theme(plot.title = element_text(size=16, face="bold.italic"))+
  labs(title = 'Historical market based poverty rate', y="Poverty rate", x="Date")+
  geom_line(size=1)

hist_urate <-  ggplot(aes(x=date, y=urate), data=natl_poverty)+
  theme_light()+
  theme(plot.title = element_text(size=16, face="bold.italic"))+
  labs(title = 'Historical unemployment 16-64', y="Unemp. rate", x="Date")+
  geom_line(size=1)


annual_natl_poverty <- fam_poverty %>% 
  group_by(year) %>% 
  summarize(n = n(),
            in_poverty = sum(inpoverty * (adj_wgt/12), na.rm=TRUE),
            poverty = weighted.mean(inpoverty, w=adj_wgt/12, na.rm=TRUE),
            urate = weighted.mean(unemp, w=adj_wgt/12, na.rm=TRUE)) %>%
  mutate(group='Poverty, all',
         poverty_earnings='All')

annual_hist_poverty <- ggplot(aes(x=year, y=poverty), data=annual_natl_poverty)+
  theme_light()+
  theme(plot.title = element_text(size=16, face="bold.italic"))+
  labs(title = 'Historical market based poverty rate', y="Poverty rate", x="Date")+
  geom_line(size=1)

hist_poverty + hist_urate

#save photo to output
ggsave(here(paste0('output/stateplots/',plotstates[i],'.png')), plot = last_plot(), dpi=300, width = 9, height = 5)


```




```{r Workbook export}
pov <- createWorkbook()

addWorksheet(pov, sheetName = "Table 1")
addWorksheet(pov, sheetName = "Table 2")

pct = createStyle(numFmt = '0.0%')
acct = createStyle(numFmt = '#.0' )
hs1 <- createStyle(fgFill = "#4F81BD", halign = "CENTER", textDecoration = "Bold",
                   border = "Bottom", fontColour = "white")

writeData(pov, headerStyle = hs1, table1, sheet = "Table 1",
          startCol = 1, startRow = 1, colNames = TRUE)
writeData(pov, table2, headerStyle = hs1, sheet = "Table 2",
          startCol = 1, startRow = 1, colNames = TRUE)

#add percent format
addStyle(pov, "Table 1", style=pct, cols=6, rows=2:(nrow(table1)+1), gridExpand=TRUE)
addStyle(pov, "Table 2", style=pct, cols=c(3:6), rows=2:(nrow(table2)+1), gridExpand=TRUE)

#add accounting format
addStyle(pov, "Table 1", style=acct, cols=c(3:5), rows=2:(nrow(table1)+1), gridExpand=TRUE)


saveWorkbook(pov, here("output/poverty_workbook.xlsx"), overwrite = TRUE)
```




